---
title       : Other indicators
subtitle    : Hartford regional data and indicators
author      : Scott Gaul
job         : Community Indicators Project
framework   : minimal        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
ext_widgets : {rCharts: ["libraries/nvd3"]} 
mode        : selfcontained # {standalone, draft}
---

## Other indicators

This page reviews several additional indicators for consideration. Generally, these indicators are used rarely, if at all, in other reports but may be of particular relevance for the Hartford region. 

### 211 calls data

United way of CT publishes some [data on calls to the 211](http://www.ctunitedway.org/media.asp). Data on call volumes by category and by town is available which would show the volume and trends in requests for assistance across the region in more detail. 

### Adult literacy

There is little to no current data on adult literacy for the Hartford region. The most recent national adult literacy survey (using the PIAAC) instrument estimated that 1 out of 6 adults has low literacy skills. NCES has [published county-level estimates](http://nces.ed.gov/naal/estimates/StateEstimates.aspx) for the 2003 and 1992 national surveys, which show a substantial improvement in literacy rates for Hartford and Tolland counties during that time.

### Chronic absenteeism

[Data to come]

### Commuting patterns

Most towns in the region have a high percentage of commuters – over 75 percent of workers in every town except Hartford. At the same time, twenty percent of all workers in the region commute to Hartford. Enfield, Manchester and Simsbury also act as local hubs for employment. The maps below show where residents of selected towns in the region travel every day for work, i.e. for each town, the map shows where the residents of that town commute to for work.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=12}
library(RColorBrewer)
ggplot(data = commutemap) + 
  geom_map(aes(map_id = id, fill = ratio), map = CTTowns) +
  geom_path(data = CTTowns, aes(x = long, y = lat, group = group),alpha = 0.05) +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL) +
  theme_minimal() +
  scale_fill_gradientn(colours = brewer.pal(4, "Blues"), name = "% of workers\n commuting to\n destination") +
  expand_limits(x = CTTowns$long, y = CTTowns$lat) +
  theme(strip.text.x = element_text(size = 12)) +
  coord_equal() + 
  facet_wrap(~ Starting.point, ncol = 4)  
```

### Public transportation usage

The Census collects data on commuting patterns and we can look at commuting patterns for the Metro Hartford region. 

For those that commute to work in Metro Hartford, driving alone is an increasingly frequent means of transport, instead of alternatives like carpooling, public transit, walking or biking. As of 2000, 83 percent of Hartford-area residents drove to work alone (compared to 67 percent in 1980), one of the highest rates in the country. Only Detroit, Grand Rapids, St. Louis, Kansas City and Indianapolis had higher rates when compared to other metro areas (as in the graphs below). (This data is old but could be updated easily.)

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 5} 
ggplot(data = walk, aes(x = Year, y = value)) + 
  geom_line(aes(group = Area), alpha = 0.15) + 
  geom_line(data = subset(walk, Area == "Hartford"), aes(x = Year, y = value), colour = "blue", size = 1) + 
  theme_minimal() +
  scale_x_continuous(breaks = c(1980,1990,2000)) +
  labs(x = NULL, y = "Percent of commuters") +
  theme(strip.text.x = element_text(size =12)) +
  facet_grid(. ~ variable)
```

### Re-entry rates

Data on re-entry rates are not currently published by the CT Department of Correction. However, we are requesting if data on re-entry rates can be shared with this group as a potential indicator. More soon...

### School enrollment patterns

School choice is one of the characteristic attributes of education in the Hartford region. While not a standard indicator as such, school enrollment patterns reveal how parents make choices about their children's education. 

The percent of school-age residents attending their local public school district can be low - less than 2/3rds in Bloomfield - and consequently other school performance data for these same districts only represents a fraction of the population. (In other words, data on Bloomfield Public Schools only describes 2/3rds of the residents of Bloomfield.)  

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 5} 
enroll$Local = ifelse(enroll$School.Type == "Public","Local public schools","Other")

enroll_agg <- ddply(subset(enroll, variable == "Total"), .(Local, Resident.Town), summarise, x = sum(x, na.rm = TRUE))
enroll_agg <- cast(enroll_agg, Resident.Town ~ Local)
names(enroll_agg) <- make.names(names(enroll_agg))
enroll_agg$Public.Ratio = enroll_agg$Local.public.schools / (
  enroll_agg$Local.public.schools + enroll_agg$Other)
enroll_agg$Resident.Town <- reorder(enroll_agg$Resident.Town, -enroll_agg$Public.Ratio)
ggplot(data = enroll_agg) +
  geom_point(aes(x = Public.Ratio, y = Resident.Town)) + 
  labs(x = "% school-age residetns enrolled in local public school district (2006-2010)", 
       y = "Resident Town") + 
  scale_x_continuous(labels = percent, limits = c(0.5,1)) +
  theme_minimal()
```

The chart below looks at the same enrollment patterns by grade. Enrollment patterns in many districts shift between pre-K / K programs and first grade, and at the transitions to middle school and high school. For example, in Bloomfield, roughly half of 10th graders living in Bloomfield don't attend Bloomfield public schools, instead opting for magnet, charter and private school options. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 12} 
library(scales)
ggplot(data = subset(enroll, School.Year == "2010-11" & variable != "Total"), 
       aes(x = variable, y = x, group = School.Type, fill = School.Type)) +
  geom_area(aes(order = as.numeric(School.Type)), position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_x_discrete(breaks = c("K","G3","G6","G9","G12")) +
  labs(x = NULL, y = "% of school-age town residents", title = "2010-11 school enrollment patterns (PK - 12)") +
  scale_fill_brewer() +
  theme_minimal() +
#  theme(strip.text.x = element_text(size = 12)) +
  facet_wrap(~ Resident.Town, ncol = 4)
```

### School languages

Given the poor test results for ELL students in the region, data on school languages can provide additional insight as to who ELL students are and where they learn. 

Here are a few factoids on foreign languages in Hartford-area schools: 
* 97 languages are spoken in schools in the region
* 3 out of 4 ELL students are Spanish-speaking
* 60 percent of ELL students attend Hartford Public Schools
* West Hartford Public Schools is the most linguistically diverse district in the region, with 46 distinct languages reported in 2010-11. (Hartford has 43, while Manchester and East Hartford both have 33.)

The visual below shows the distribution of ELL students by district and the concentration of particular languages in the region (for any occurence of more than 5 students speaking a particular language in a district). Outside of Spanish, there are significant groups speaking Vietnamese, Chinese, Serbo-Croatian, Urdu, Portuguese, Arabic, and Karen (from Myanmar), among others. 

[graph pending]

### School discipline / arrests

[Data to come]
